### 2024-09-24 paper/paper.qmd
add some graph, introduction, abstract in the paper change title of readme change simulated data statistics

---
title: "Analysis on Trend of Death of Shelter Residence in Toronto "
subtitle: "focus on gender distribution from 2007 to 2024"
author: 
  - Xinqi Yue
thanks: "Code and data are available at: https://github.com/xinqiyue/ShelterDeath."
date: today
date-format: long
abstract: "Trough the analysis on the data about deaths of shelter residents, we found that the total number of deaths of shelter residences is increasing year by year. The proportion of male deaths is relatively higher. The study will help understanding the dynamic of mortality and making better policy and intervine to avoid similar problems in the future."
format: pdf
number-sections: true
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

The rising number of deaths within Toronto's shelter system is a growing area of concern, reflecting broader societal challenges such as homelessness, mental health crises, and inadequate access to health care. Understanding the dynamics of these deaths is critical to developing effective policies and interventions aimed at reducing mortality in this vulnerable population. This paper focuses on analyzing the dataset related to deaths in Toronto's shelter residence, and seeks to identify the demographic trends and factors contributing to these unfortunate outcomes.

Despite the importance of this issue, there is a clear gap in the public's detailed understanding of who is being affected and how patterns of mortality are evolving over time. Due to privacy concerns, the dataset is limited in scope and lacks identifying information such as names, dates of birth, and specific causes of death. Also, the Non-binary genders are not included until 2020. Nevertheless, these aggregate data allow for an examination of key trends, such as the gender distribution of deaths and the change over time. These trends highlight critical disparities, particularly the higher proportion of male deaths and the steady increase in overall mortality number.

This study aims to provide a visual and statistical analysis of the available data and to draw attention to the factors that may be contributing to the increased mortality. The analysis reveals two important findings: first, the number of deaths has steadily increased over the years, indicating a deteriorating situation within the shelter system; Second, men make up a larger proportion of the deceased, suggesting potential gender-specific vulnerabilities.

These findings are important because they provide a clearer picture of the extent and nature of mortality in Toronto's shelter population, which can be used to inform future policy decisions and the allocation of resources. By identifying key trends and limitations within the dataset, this study provides a basis for more targeted and informed interventions to address the underlying causes of death in this population. The structure of this paper begins with a detailed description of the data@sec-data, followed by the statistic model@sec-data, results@sec-data, and discussion of implications@sec-data.

# Data {#sec-data}

The dataset used for this analysis relates to deaths within Toronto's shelter system. It contains key demographic variables and aggregates information about the number of deaths reported each year, categorized by gender. However, due to privacy concerns and respect for the confidentiality of individuals who have passed, the dataset excludes personal identifying information including names, birthdates, or specific locations of death. This limitation necessitated the use of data for analysis focusing on factors including gender distribution and time rather than locations and age.

##Variables
The dataset include the following variables:

Year: The year in which the death occurred at a Toronto shelter, ranging from 2007 to 2024. This variable aids in analyzing trends over time.
Month: The month in which the death occurred, corresponding to each year from January to December. Data beyond August 2024 has not yet been recorded.
Total Deaths: The total number of deaths recorded in the given month.
Male Deaths: The total number of male shelter residents who died in the given month.
Female Deaths: The total number of female shelter residents who died in the given month.
Other Deaths: The number of deaths of individuals identified as Transgender, Non-binary, or Two-Spirit in the given month. Note that data for this category was not recorded prior to 2020.

Some of our data is of penguins (@fig-bills), from

```{r fig-annual-death, echo=FALSE, message=FALSE, warning=FALSE}
shelter_death_data <- read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv", show_col_types = FALSE)

# calculate the annual death
annual_deaths <- shelter_death_data %>%
  filter(year >= 2007 & year <= 2023) %>%
  group_by(year) %>%
  summarise(total_yearly_deaths = sum(total_decedents), na.rm = TRUE)

# get line graph, showing the annual death trend
ggplot(annual_deaths, aes(x = year, y = total_yearly_deaths)) +
  geom_line(size = 1, color = "black") +  # get line graph
  geom_point(size = 3, color = "pink") +  # add points on value in each year
  labs(title = "Total Deaths in Toronto Shelters (2007-2023)",
       x = "Year",
       y = "Total Deaths") +  # as title and name of the axis
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

```{r fig-monthly-death, echo=FALSE, message=FALSE, warning=FALSE}

# calculate the annual death
month_deaths <- shelter_death_data %>%
  filter(year >= 2007 & year <= 2023) %>%
  group_by(month) %>%
  summarise(total_monthly_deaths = sum(total_decedents))

month_deaths$month <- factor(month_deaths$month, levels = month.abb)

# get line graph, showing the annual death trend
ggplot(month_deaths, aes(x = month, y = total_monthly_deaths, group = 1)) +
  geom_line(size = 1, color = "black") +  # get line graph
  geom_point(size = 3, color = "pink") +  # add points on value in each year
  labs(title = "Total Deaths in Toronto Shelters (2007-2023)",
       x = "Month",
       y = "Total Deaths") +  # as title and name of the axis
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
```

```{r fig-shelter-death1, echo=FALSE, message=FALSE, warning=FALSE}

shelter_death_data07to15 <- shelter_death_data %>%
  filter(year >= 2007 & year <= 2015)

shelter_death_data15to24 <- shelter_death_data %>%
  filter(year >= 2016)

# get line graph, showing the annual death trend
ggplot(shelter_death_data07to15, aes(x = month, y = total_decedents, group = year)) +
  geom_line(size = 1, color = "black") +  # get line graph
  geom_point(size = 1, color = "pink") +  # add points on value in each year
  labs(title = "Total Deaths in Toronto Shelters (2007-2015)",
       x = "Month",
       y = "Total Deaths") +  # as title and name of the axis
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title.y = element_text(size = 12),  # 只显示y轴的标签
    axis.text.x = element_blank(),  # 移除x轴刻度
    axis.ticks.x = element_blank(),  # 移除x轴刻度线
    panel.spacing = unit(1, "lines")  # 增加子图间的间隔
  ) +
  facet_wrap(~ year, ncol = 3)
```

```{r fig-shelter-death2, echo=FALSE, message=FALSE, warning=FALSE}
# get line graph, showing the annual death trend
ggplot(shelter_death_data15to24, aes(x = month, y = total_decedents, group = year)) +
  geom_line(size = 1, color = "black") +  # get line graph
  geom_point(size = 1, color = "pink") +  # add points on value in each year
  labs(title = "Total Deaths in Toronto Shelters (2015-2024)",
       x = "Month",
       y = "Total Deaths") +  # as title and name of the axis
  theme_minimal() + 
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    axis.title.y = element_text(size = 12),  # 只显示y轴的标签
    axis.text.x = element_blank(),  # 移除x轴刻度
    axis.ticks.x = element_blank(),  # 移除x轴刻度线
    panel.spacing = unit(1, "lines")  # 增加子图间的间隔
  ) +
  facet_wrap(~ year, ncol = 3)

```

Talk more about it.

And also planes (@fig-planes). (You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)

```{r}

```

Talk way more about it.

# Model{#sec-model}

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.

\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}

We run the model in R [@citeR] using the `rstanarm` package of . We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance $\theta$.

# Results{#sec-result}

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

```

# Discussion{#sec-discussion}

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.

## Second discussion point

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {-}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

```

\newpage

# References

### 2024-09-24 scripts/00-simulate_data.R

@@ -26,7 +26,7 @@ years <- format(dates, "%Y")
 
 months <- format(dates, "%b")
 
-simulated_shelter_death <- rpois(length(months),lambda =  1)
+simulated_shelter_death <- rpois(length(months),lambda =  3)
 
 male_death <- round(simulated_shelter_death * 0.65)
 

### 2024-09-24 paper/paper.qmd
add llm usage relocate sketch add limitations in paper and more graphs(bar charts)

@@
 The proportion of male deaths is relatively higher.
-The study will help understanding the dynamic of mortality and making better policy and intervine to avoid similar problems in the future."
+The findings provide insights into the factors affecting shelter residents' well-being and can guide policies to address homelessness, improve living conditions, and tackle health-related issues in this vulnerable population."
 format: pdf
@@
 By identifying key trends and limitations within the dataset, this study provides a basis for more targeted and informed interventions to address the underlying causes of death in this population.
-The structure of this paper begins with a detailed description of the data@sec-data, followed by the statistic model@sec-data, results@sec-data, and discussion of implications@sec-data.
+The structure of this paper begins with a detailed description of the data@sec-data, followed by the statistic model@sec-model, results@sec-result, and discussion of implications@sec-discussion.
 
 # Data {#sec-data}
-
 The dataset used for this analysis relates to deaths within Toronto's shelter system.
@@
 
-##Variables
+## Variables
 The dataset include the following variables:
 
-Year: The year in which the death occurred at a Toronto shelter, ranging from 2007 to 2024. This variable aids in analyzing trends over time.
-Month: The month in which the death occurred, corresponding to each year from January to December. Data beyond August 2024 has not yet been recorded.
-Total Deaths: The total number of deaths recorded in the given month.
-Male Deaths: The total number of male shelter residents who died in the given month.
-Female Deaths: The total number of female shelter residents who died in the given month.
-Other Deaths: The number of deaths of individuals identified as Transgender, Non-binary, or Two-Spirit in the given month. Note that data for this category was not recorded prior to 2020.
+1. **Year**: The year in which the death occurred at a Toronto shelter, ranging from 2007 to 2024. This variable aids in analyzing trends over time.
 
-Some of our data is of penguins (@fig-bills), from
+2. **Month**: The month in which the death occurred, corresponding to each year from January to December. Data beyond August 2024 has not yet been recorded.
+
+3. **Total Deaths**: The total number of deaths recorded in the given month.
+
+4. **Male Deaths**: The total number of male shelter residents who died in the given month.
+
+5. **Female Deaths**: The total number of female shelter residents who died in the given month.
+
+6. **Other Deaths**: The number of deaths of individuals identified as Transgender, Non-binary, or Two-Spirit in the given month. Note that data for this category was not recorded prior to 2020.
+
+## Limitation
+The dataset used for this analysis presents several limitations that must be acknowledged to provide a comprehensive understanding of the findings.
+
+**Privacy and Respect for the Deceased**: Due to ethical considerations, certain sensitive information such as age and location of death was not collected.
+This lack of data may limit the ability to perform more granular analyses related to demographic factors influencing mortality rates among shelter residents.
+
+**Gender Data Constraints**: Information on gender identity, specifically regarding transgender, non-binary, and two-spirit individuals, was not systematically recorded until 2020.
+As a result, prior to this year, any data related to these gender identities is missing.
+This gap restricts the ability to fully analyze trends and disparities in mortality rates among diverse gender identities over time.
+
+**Absence of Comparable Datasets**: It is essential to consider that the dataset could be enriched by including broader contextual variables.
+Specifically, information about the total shelter population and its gender composition, age distribution, and health status of residents would provide a more nuanced understanding of the factors contributing to mortality in these settings.
+Such data would allow for deeper insights into the demographic profiles of those affected and could better inform policies aimed at addressing the challenges of homelessness and shelter living.
+
+**Data Quality and Cleaning**: The dataset required careful cleaning to address issues such as missing values and inconsistencies, particularly in the reporting of deaths by gender.
+Although high-level cleaning was conducted to ensure the reliability of the data, the absence of certain variables may still introduce biases or limit the interpretability of the results.
+
+By recognizing these limitations, the analysis aims to provide a clearer context for the findings and to encourage further research that can fill these gaps and enhance understanding of mortality trends among shelter residents.
+
+## Similar Datasets
+
+## Overall Visualization
+
+### Trend Over Time
+
+Figure 1 shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year.
+This layout allows readers to clearly compare the trend of deaths from one year to the next and observe potential seasonal changes.
+
+```{r fig-shelter-death-line, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
+# Read the analysis data of shelter residence death.
+shelter_death_data <- read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv", show_col_types = FALSE)
+
+# Sort the month data in order
+shelter_death_data$month <- factor(shelter_death_data$month, levels = month.abb)
+
+# Get line graph, showing the death trend 
+ggplot(shelter_death_data, aes(x = month, y = total_decedents, group = year)) +
+  geom_line(size = 1, color = "black") +  # get line graph
+  geom_point(size = 2, color = "pink") +  # add points on value in each year
+  labs(title = "Total Deaths in Toronto Shelters (2007-2024)",
+       x = "Month",
+       y = "Total Deaths") +  # add title and name of the axis
+  theme_minimal() + 
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate horizontal axis labels to avoid overlapping
+  ) +
+  facet_wrap(~ year, ncol = 3)
+
+```
+Based on the generated line graphs, the following conclusions can be drawn:
+
+1.
+**Overall Increasing Trend in Deaths**: From 2007 to 2024, there has been a consistent increase in the number of deaths.
+This trend may be related to several factors, including changes in economic conditions, the impact of the COVID-19 pandemic, and overall population growth.
+These factors could contribute to a rise in the homeless population, leading to an increase in mortality rates.
+
+2.
+**Peak Deaths in October 2021**: Notably, October 2021 recorded the highest number of deaths.
+This peak may be associated with the public health crisis stemming from the COVID-19 pandemic, social isolation policies, and their effects on mental health and living conditions.
+Further research could help clarify the specific contributing factors during this period.
+
+3.
+**Lack of Significant Seasonal Impact on Deaths**: Overall, there appears to be no substantial seasonal variation in death rates, indicating that the fluctuations in mortality may be more closely linked to economic and social factors rather than seasonal influences.
+
+These conclusions provide important insights into the overall mortality trends of Toronto shelter residents and suggest that relevant agencies need to consider multiple influencing factors when developing policies and interventions.
+
+### Gender Composition
+Figure 2
+
+```{r fig-shelter-death-bar, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
+annual_death_counts_bar <- shelter_death_data %>%
+  filter(year >= 2007 & year <= 2023) %>%
+  group_by(year) %>%
+  summarise(
+    total_deaths = sum(total_decedents),
+    male_deaths = sum(male),
+    female_deaths = sum(female),
+    other_deaths = sum(transgender_non_binary_two_spirit, na.rm = TRUE)
+  ) %>%
+  pivot_longer(cols = c(male_deaths, female_deaths, other_deaths), 
+               names_to = "gender", 
+               values_to = "count")
+
+# Create bar chart showing annual death counts in Toronto shelters from 2007 to. 2023
+ggplot(annual_death_counts_bar, aes(x = factor(year), y = count, fill = gender)) +
+  geom_bar(stat = "identity", position = "stack") +
+  labs(title = "Annual Death Counts in Toronto Shelters (2007-2023)",
+       x = "year",
+       y = "num of death",
+       fill = "gender") +
+  scale_fill_manual(values = c("male_deaths" = "pink", "female_deaths" = "black", "other_deaths" = "gray"), labels = c("Female", "male", "Other")) +  # change color and the text on the side
+  theme_minimal() + 
+  theme(
+    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
+    axis.title = element_text(size = 12),
+    axis.text = element_text(size = 10),
+    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate horizontal axis labels to avoid overlapping
+  )
+```
+
+### Compariasion Over Year
 
 ```{r fig-annual-death, echo=FALSE, message=FALSE, warning=FALSE}
-shelter_death_data <- read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv", show_col_types = FALSE)
 
@@
        x = "Year",
-       y = "Total Deaths") +  # as title and name of the axis
+       y = "Total Deaths") +  # add title and name of the axis
   theme_minimal() + 
@@
 ```
+
+### Compariasion Over Month
 
@@
 ```
-
-```{r fig-shelter-death1, echo=FALSE, message=FALSE, warning=FALSE}
-
-shelter_death_data07to15 <- shelter_death_data %>%
-  filter(year >= 2007 & year <= 2015)
-
-shelter_death_data15to24 <- shelter_death_data %>%
-  filter(year >= 2016)
-
-# get line graph, showing the annual death trend
-ggplot(shelter_death_data07to15, aes(x = month, y = total_decedents, group = year)) +
-  geom_line(size = 1, color = "black") +  # get line graph
-  geom_point(size = 1, color = "pink") +  # add points on value in each year
-  labs(title = "Total Deaths in Toronto Shelters (2007-2015)",
-       x = "Month",
-       y = "Total Deaths") +  # as title and name of the axis
-  theme_minimal() + 
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
-    axis.title = element_text(size = 12),
-    axis.text = element_text(size = 10),
-    axis.title.y = element_text(size = 12),  # 只显示y轴的标签
-    axis.text.x = element_blank(),  # 移除x轴刻度
-    axis.ticks.x = element_blank(),  # 移除x轴刻度线
-    panel.spacing = unit(1, "lines")  # 增加子图间的间隔
-  ) +
-  facet_wrap(~ year, ncol = 3)
-```
-
-```{r fig-shelter-death2, echo=FALSE, message=FALSE, warning=FALSE}
-# get line graph, showing the annual death trend
-ggplot(shelter_death_data15to24, aes(x = month, y = total_decedents, group = year)) +
-  geom_line(size = 1, color = "black") +  # get line graph
-  geom_point(size = 1, color = "pink") +  # add points on value in each year
-  labs(title = "Total Deaths in Toronto Shelters (2015-2024)",
-       x = "Month",
-       y = "Total Deaths") +  # as title and name of the axis
-  theme_minimal() + 
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
-    axis.title = element_text(size = 12),
-    axis.text = element_text(size = 10),
-    axis.title.y = element_text(size = 12),  # 只显示y轴的标签
-    axis.text.x = element_blank(),  # 移除x轴刻度
-    axis.ticks.x = element_blank(),  # 移除x轴刻度线
-    panel.spacing = unit(1, "lines")  # 增加子图间的间隔
-  ) +
-  facet_wrap(~ year, ncol = 3)
-
-```
-
-Talk more about it.
-
-And also planes (@fig-planes).
-(You can change the height and width, but don't worry about doing that until you have finished every other aspect of the paper - Quarto will try to make it look nice and the defaults usually work well once you have enough text.)
 

### 2024-09-24 scripts/02-data_cleaning.R

@@ -18,6 +18,29 @@ cleaned_data <-
   janitor::clean_names() |>
   tidyr::drop_na()
 
+# Calculate the proportion of male in total shelter deaths
+cleaned_data
+cleaned_data <- cleaned_data %>%
+  mutate(male_ratio = male / total_decedents)
+
+# Calculate the proportion of female in total shelter deaths
+cleaned_data <- cleaned_data %>%
+  mutate(female_ratio = female / total_decedents)
+
+# Replace "n/a" with NA
+cleaned_data <- cleaned_data %>%
+  mutate(transgender_non_binary_two_spirit = ifelse(transgender_non_binary_two_spirit == "n/a", NA, transgender_non_binary_two_spirit))
+
+# Transfer transgender_non_binary_two_spirit death num to numeric data
+cleaned_data <- cleaned_data %>%
+  mutate(transgender_non_binary_two_spirit = as.numeric(transgender_non_binary_two_spirit))
+
+
+# Calculate the proportion of transgender, non_binary or two spirit in total shelter deaths
+cleaned_data <- cleaned_data %>%
+  mutate(other_ratio = ifelse(is.na(transgender_non_binary_two_spirit) | total_decedents == 0, NA, transgender_non_binary_two_spirit / total_decedents))
+
+
 #### Save data ####
 write_csv(cleaned_data, "data/analysis_data/shelter_residences_death_analysis_data.csv") 
 

### 2024-09-24 paper/paper.qmd
Add more citations, bib and llm usage in the paper. Add more text content to illustrate the graphs.

@@
 By identifying key trends and limitations within the dataset, this study provides a basis for more targeted and informed interventions to address the underlying causes of death in this population.
-The structure of this paper begins with a detailed description of the data@sec-data, followed by the statistic model@sec-model, results@sec-result, and discussion of implications@sec-discussion.
+The structure of this paper begins with a detailed description of the data @sec-data, followed by the statistic results @sec-result, and discussion of implications @sec-discussion.
 
 # Data {#sec-data}
-The dataset used for this analysis relates to deaths within Toronto's shelter system.
+The dataset used for this analysis relates to deaths within Toronto's shelter system, from @toronto_shelter_deaths.
 It contains key demographic variables and aggregates information about the number of deaths reported each year, categorized by gender.
@@
 ## Similar Datasets
+The dataset titled Deaths of People Experiencing Homelessness, from @toronto_homelessness_deaths, provides information on the deaths of homeless individuals, which is somewhat similar to our current study of deaths among Toronto shelter residents.
+One notable advantage of this dataset is that it includes a detailed CSV file outlining the causes of death, offering valuable insights into the factors contributing to mortality in this population.
+
+However, the dataset containing the causes of death cannot be downloaded directly, which limits the accessibility and availability of our analysis.
+The accessible data variables align perfectly with our research dataset but only cover the period from 2017 to 2023.
+The lack of longitudinal data means it does not align with the comprehensive timeline of our study, which spans from January 2007 to August 2024.
+Therefore, although this dataset provides certain advantages, the limitations regarding its time span and accessibility led to its exclusion from our viable research options.
 
 ## Overall Visualization
+In the following section, I will present a series of visualizations that analyze the number of deaths among shelter residents over time, highlighting the gender distribution and overall data trends.
+These charts will provide a clearer understanding of the data composition and allow for an insightful examination of how the number of deaths in this population has evolved.
 
@@
 
-Figure 1 shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year.
+Figure @fig-shelter-death-line, created by @citeR, shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year.
 This layout allows readers to clearly compare the trend of deaths from one year to the next and observe potential seasonal changes.
 
-```{r fig-shelter-death-line, echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
+```{r fig-shelter-death-line, fig.cap="Deaths in Toronto Shelters Over Time (2007.01-2024.08)", echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
 # Read the analysis data of shelter residence death.
@@
   geom_point(size = 2, color = "pink") +  # add points on value in each year
-  labs(title = "Total Deaths in Toronto Shelters (2007-2024)",
+  labs(title = "Deaths in Toronto Shelters Over Time (2007.01-2024.08)",
        x = "Month",
@@
 
-### Gender Composition
-Figure 2
+### Annual Deaths with Gender Composition
+Figure @fig-annual-death, created by @citeR, below illustrates the annual number of deaths from 2007 to 2023.
+And Figure @fig-shelter-death-bar, created by @citeR, illustrates the annual number of deaths of the same time period, with distinct color coding to represent gender distribution: pink for male, black for female, and gray for other gender identities, including transgender and non-binary.
 
-```{r fig-shelter-death-bar, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
+Several important conclusions can be drawn from the visualization.
+First, the proportion of male deaths remains consistently the highest throughout the years observed, indicating a significant gender disparity in mortality rates among shelter residents.
+In addition, there is a noticeable upward trend in the overall number of deaths, suggesting a growing concern for the well-being of this vulnerable population.
+It can also be seen clearly from the line graph of Figure @fig-annual-death.
+Notably, 2021 stands out as the peak year for mortality.
+This can be seen clearly in Figure @fig-annual-death too.
+This raises questions about the factors contributing to this spike and the broader implications for public health and social support systems.
+
+```{r fig-shelter-death-bar, fig.cap="Annual Death Counts in Toronto Shelters With Gender Distribution (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
 annual_death_counts_bar <- shelter_death_data %>%
@@
 
-# Create bar chart showing annual death counts in Toronto shelters from 2007 to. 2023
+# Create bar chart showing annual death counts in Toronto shelters from 2007 to 2023.
 ggplot(annual_death_counts_bar, aes(x = factor(year), y = count, fill = gender)) +
   geom_bar(stat = "identity", position = "stack") +
-  labs(title = "Annual Death Counts in Toronto Shelters (2007-2023)",
+  labs(title = "Annual Death Counts in Toronto Shelters With Gender Distribution (2007-2023)",
        x = "year",
@@
 
-### Compariasion Over Year
-
-```{r fig-annual-death, echo=FALSE, message=FALSE, warning=FALSE}
+```{r fig-annual-death, fig.cap="Annual Deaths in Toronto Shelters (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
 
@@
   geom_point(size = 3, color = "pink") +  # add points on value in each year
-  labs(title = "Total Deaths in Toronto Shelters (2007-2023)",
+  labs(title = "Annual Deaths in Toronto Shelters (2007-2023)",
        x = "Year",
@@
 ```
-
-```{r}
-
-```
-
-Talk way more about it.
-
-# Model{#sec-model}
-
-The goal of our modelling strategy is twofold. Firstly,...
-
-Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].
-
-## Model set-up
-
-Define $y_i$ as the number of seconds that the plane remained aloft. Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both measured in millimeters.
-
-\begin{align} 
-y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
-\mu_i &= \alpha + \beta_i + \gamma_i\\
-\alpha &\sim \mbox{Normal}(0, 2.5) \\
-\beta &\sim \mbox{Normal}(0, 2.5) \\
-\gamma &\sim \mbox{Normal}(0, 2.5) \\
-\sigma &\sim \mbox{Exponential}(1)
-\end{align}
-
-We run the model in R [@citeR] using the `rstanarm` package of . We use the default priors from `rstanarm`.
-
-### Model justification
-
-We expect a positive relationship between the size of the wings and time spent aloft. In particular...
-
-We can use maths by including latex between dollar signs, for instance $\theta$.
 
@@
 
-# Model details {#sec-model-details}
-
-## Posterior predictive check
-
-In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive check. This shows...
-
-In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the prior. This shows...
-
-```{r}
-#| eval: true
-#| echo: false
-#| message: false
-#| warning: false
-#| label: fig-ppcheckandposteriorvsprior
-#| layout-ncol: 2
-#| fig-cap: "Examining how the model fits, and is affected by, the data"
-#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]
-
-```
-
-## Diagnostics
-
-@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...
-
-@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...
-
-```{r}
-#| echo: false
-#| eval: true
-#| message: false
-#| warning: false
-#| label: fig-stanareyouokay
-#| fig-cap: "Checking the convergence of the MCMC algorithm"
-#| fig-subcap: ["Trace plot", "Rhat plot"]
-#| layout-ncol: 2
-
-```
-
 \newpage

### 2024-09-24 paper/paper.qmd
add illustration of trend over month graph

@@
   theme(
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
-    axis.title = element_text(size = 12),
-    axis.text = element_text(size = 10)
+    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
+    axis.title = element_text(size = 10),
+    axis.text = element_text(size = 8)
   )
@@
 
+\newpage
+
 ### Compariasion Over Month
+Figure @fig-monthly-death shows the aggregated monthly death counts for shelter residents from 2007 to 2023.
+This data was derived by summing the death counts for each January from 2007 to 2023, resulting in a comprehensive data set for January, and repeating this process for each of the other months, resulting in data for all twelve months.
 
-```{r fig-monthly-death, echo=FALSE, message=FALSE, warning=FALSE}
+The purpose of this visualization is to analyze whether certain months have higher mortality rates.
+It opens the door to exploring potential seasonal influences on mortality rates, such as the impact of extreme temperatures or adverse weather conditions, which could contribute to increased mortality among shelter residents.
+
+The current findings, shown in the line graph, indicate that January and November have the highest mortality rates, while March and September have the lowest mortality rates.
+These findings highlight the need for further research into the factors influencing these monthly variations, which could inform targeted interventions to improve the health and safety of shelter residents.
+
+```{r fig-monthly-death, fig.cap="Monthly Mortality Trends Among Shelter Residents (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
 
@@
   geom_point(size = 3, color = "pink") +  # add points on value in each year
-  labs(title = "Total Deaths in Toronto Shelters (2007-2023)",
+  labs(title = "Monthly Mortality Trends Among Shelter Residents (2007-2023)",
        x = "Month",

### 2024-09-25 paper/paper.qmd
delete unnecessary files add content in discussion and result remove unnecessay new column in analysis data

@@
 
+library(kableExtra)
 library(tidyverse)
 library(palmerpenguins)
+
+# Read the analysis data of shelter residence death.
+shelter_death_data <- read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv", show_col_types = FALSE) 
 ```
@@
 By identifying key trends and limitations within the dataset, this study provides a basis for more targeted and informed interventions to address the underlying causes of death in this population.
-The structure of this paper begins with a detailed description of the data @sec-data, followed by the statistic results @sec-result, and discussion of implications @sec-discussion.
+This paper begins with a comprehensive overview of the dataset, including its variables (@sec-var), limitations (@sec-lim), the rationale for its selection (@sec-sim), and the data collection process (@sec-measurement).
+Following this, the data is presented with accompanying graphs and analysis (@sec-overview).
+The subsequent section outlines key findings (@sec-result), followed by a discussion of their broader implications (@sec-discussion).
+All visualizations were generated using @citeR, with the data sourced from the @toronto_shelter_deaths dataset.
 
@@
 
-## Variables
+## Variables {#sec-var}
 The dataset include the following variables:
@@
 
-## Limitation
+The variable follows the rule:
+Total Death = Male Deaths + Female Deaths + Other Deaths
+
+## Limitation {#sec-lim}
 The dataset used for this analysis presents several limitations that must be acknowledged to provide a comprehensive understanding of the findings.
@@
 
-## Similar Datasets
+## Similar Datasets {#sec-sim}
 The dataset titled Deaths of People Experiencing Homelessness, from @toronto_homelessness_deaths, provides information on the deaths of homeless individuals, which is somewhat similar to our current study of deaths among Toronto shelter residents.
@@
 
-## Overall Visualization
+## Measurement {#sec-measurement}
+Within this data set, the Structured Reporting System facilitates the collection of information regarding the death of shelter residents.
+Shelter operators are required to notify the City of Toronto within 24 hours of the death of a current or recent shelter resident and then submit a detailed written report within 30 days.
+This systematic approach, in place since 2007 and reviewed annually by Toronto Housing and Support Services, ensures timely and consistent data collection.
+The structured reporting requirement imposes a level of discipline on shelter operators to ensure that deaths are not only reported in a timely manner, but also fully documented, thereby enhancing the integrity and credibility of the dataset.
+
+ However, certain limitations affect the completeness and reliability of the data.
+For example, due to privacy concerns, personal information such as names, dates of birth, and specific places of death are not collected.
+This limits the granularity of the data and can hinder deeper analysis, such as correlations with demographic or location-specific factors.
+In addition, shelter operators often do not have access to death certificates or information on causes of death, so the dataset lacks this critical context that could be important for understanding the factors contributing to mortality among shelter residents.
+
+Improvements to the measurement process could include better integration with public health and legal entities to access more detailed information about deaths, including cause of death, while still protecting individual privacy.
+This improvement would provide a more holistic view of the factors influencing mortality rates among shelter residents, allowing for more informed policy decisions to address health and safety issues in shelters.
+
+## Overall Visualization {#sec-overview}
 In the following section, I will present a series of visualizations that analyze the number of deaths among shelter residents over time, highlighting the gender distribution and overall data trends.
@@
 ```{r fig-shelter-death-line, fig.cap="Deaths in Toronto Shelters Over Time (2007.01-2024.08)", echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
-# Read the analysis data of shelter residence death.
-shelter_death_data <- read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv", show_col_types = FALSE)
-
 # Sort the month data in order
@@
 # Results{#sec-result}
+As time progresses, it is evident from Figures @fig-shelter-death-bar and @fig-annual-death that the number of deaths among shelter residents has increased.
+Potential reasons for this trend may include changes in the economic landscape, shifts in the overall population, policy modifications, and factors related to the pandemic.
 
-Our results are summarized in @tbl-modelresults.
+Regarding monthly variations, while Figure @fig-monthly-death illustrates a noticeable peak in January and a low in September, a comprehensive view of the data in Figure @fig-shelter-death-line suggests that seasonal changes do not appear to have a significant correlation with the number of deaths.
+One possible explanation for this phenomenon could be related to the timing of the pandemic's outbreak.
 
-```{r}
-#| echo: false
-#| eval: true
-#| warning: false
-#| message: false
+In terms of gender composition, male deaths have consistently constituted the highest proportion from 2007 to 2023, indicating a potential underlying vulnerability.
+However, this trend may also be influenced by the higher proportion of males residing in shelters.
+
+# Discussion{#sec-discussion}
+
+## suammry
+This article examines trends in mortality among shelter residents in Toronto from 2007 to 2023, with a focus on gender distribution.
+Our analysis shows that there was an upward trend in deaths during this period, with a higher proportion of men dying.
+## Weaknesses and next steps
+
+In exploring the implications of these findings, we gain valuable insights into the complex interplay of economic, social, and health factors that affect shelter residents.
+The data suggest that economic downturns, policy changes, and global health events may exacerbate the vulnerability of this population, warranting a reassessment of current support systems and interventions.
+In addition, the lack of a clear seasonal pattern in mortality suggests that other factors, such as the pandemic, may be masking typical seasonal effects and warrant further investigation.
+
+While the results highlight important issues, certain limitations must be acknowledged.
+The dataset used for this analysis may not fully capture the demographics of the shelter population, such as sex composition and health status.
+In addition, the analysis relied on reported deaths, which may be inaccurate or underreported.
+Future research should aim to incorporate more comprehensive demographic data, such as data on the total shelter population and its gender composition, causes of death, etc., to improve our understanding of the factors driving these trends.
+
+Finally, this study highlights several areas for future research.
+Further research could focus on the specific causes of mortality among shelter residents and the effectiveness of current policies and programs designed to reduce mortality among this vulnerable population.
+By addressing these gaps, we can better inform policy decisions and improve outcomes for people experiencing homelessness.
+
+\appendix
+# Appendix {-}
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+# Get first 10 lines in the data set as a sample
+example_table <- shelter_death_data %>%
+  slice(1:10)
+
+example_table %>%
+  kbl(caption = "Sample Table of Deaths in Toronto Shelters (2007-2023)",
+      col.names = c("ID", "Year", "Month", "Total decedents", "male", "female", "Others")) %>%
+  kable_styling(full_width = FALSE, position = "center", latex_options = "scale_down") %>%
+  kable_classic(full_width = FALSE, html_font = "Cambria")
 
@@
 
-```{r}
-#| echo: false
-#| eval: true
-#| label: tbl-modelresults
-#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
-#| warning: false
-
+```{r echo=FALSE, message=FALSE, warning=FALSE}
+# summarize the data group by year
+summary_table <- shelter_death_data %>%
+  filter(year >= 2007 & year <= 2023) %>%
+  group_by(year) %>%
+  summarise(
+    total_deaths = sum(total_decedents),  # total death num
+    mean_deaths = round(mean(total_decedents), 2),  # average death num each month, rounded to 2 decimal places
+    median_deaths = round(median(total_decedents), 2),  # median, rounded
+    min_deaths = min(total_decedents),  # minimum value
+    max_deaths = max(total_decedents),  # maximum value
+    male_ratio = round(sum(male) / total_deaths * 100, 2),  # ratio of male death, rounded
+    female_ratio = round(sum(female) / total_deaths * 100, 2),  # ratio of female death, rounded
+    # other_ratio = round(sum(transgender_non_binary_two_spirit, na.rm = TRUE) / total_deaths * 100, 2)  # ratio of death of other genders, rounded
+    other_ratio = ifelse(any(is.na(shelter_death_data$transgender_non_binary_two_spirit)), NA,round(sum(transgender_non_binary_two_spirit, na.rm = TRUE) / total_deaths * 100, 2))  
+    # ratio of death of other genders, rounded
+  )
+summary_table %>% 
+  kbl(caption = "Annual Summary of Deaths in Toronto Shelters (2007-2023)", 
+      col.names = c("Year", "Total Deaths", "Mean", "Median", "Min", "Max", "Male Ratio (%)", "Female Ratio (%)", "Other Ratio (%)")) %>%
+  kable_styling(full_width = FALSE, position = "center", latex_options = "scale_down") %>%
+  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
+  add_header_above(c(" " = 1, "Deaths Summary" = 5, "Gender Ratios" = 3))
 ```
-
-# Discussion{#sec-discussion}
-
-## First discussion point {#sec-first-point}
-
-If my paper were 10 pages, then should be be at least 2.5 pages. The discussion is a chance to show off what you know and what you learnt from all this.
-
-## Second discussion point
-
-## Third discussion point
-
-## Weaknesses and next steps
-
-Weaknesses and next steps should also be included.
-
-\newpage
-
-\appendix
-
-# Appendix {-}
-
-# Additional data details
 

### 2024-09-25 scripts/02-data_cleaning.R

@@ -10,35 +10,18 @@
 #### Workspace setup ####
 library(tidyverse)
 
-#### Clean data ####
+#### Get data ####
 raw_data <- read_csv("data/raw_data/shelter_residences_death_raw_data.csv") 
 
+#### Clean data ####
 cleaned_data <-
   raw_data |>
   janitor::clean_names() |>
   tidyr::drop_na()
 
-# Calculate the proportion of male in total shelter deaths
-cleaned_data
-cleaned_data <- cleaned_data %>%
-  mutate(male_ratio = male / total_decedents)
 
-# Calculate the proportion of female in total shelter deaths
-cleaned_data <- cleaned_data %>%
-  mutate(female_ratio = female / total_decedents)
-
-# Replace "n/a" with NA
 cleaned_data <- cleaned_data %>%
   mutate(transgender_non_binary_two_spirit = ifelse(transgender_non_binary_two_spirit == "n/a", NA, transgender_non_binary_two_spirit))
-
-# Transfer transgender_non_binary_two_spirit death num to numeric data
-cleaned_data <- cleaned_data %>%
-  mutate(transgender_non_binary_two_spirit = as.numeric(transgender_non_binary_two_spirit))
-
-
-# Calculate the proportion of transgender, non_binary or two spirit in total shelter deaths
-cleaned_data <- cleaned_data %>%
-  mutate(other_ratio = ifelse(is.na(transgender_non_binary_two_spirit) | total_decedents == 0, NA, transgender_non_binary_two_spirit / total_decedents))
 
 
 #### Save data ####

### 2024-09-25 paper/paper.qmd
delete model folder add overview and llm usage statement in Readme fix cross-reference and small title

@@
 The proportion of male deaths is relatively higher.
+And seasonal factors does not affect much on it.
 The findings provide insights into the factors affecting shelter residents' well-being and can guide policies to address homelessness, improve living conditions, and tackle health-related issues in this vulnerable population."
@@
 These charts will provide a clearer understanding of the data composition and allow for an insightful examination of how the number of deaths in this population has evolved.
+There are 212 observations in the original dataset in total and there is no special cleaning methods needed as the data is already clear enough.
+See Appendix @Appendix for the summary of the dataset.
 
@@
 Our analysis shows that there was an upward trend in deaths during this period, with a higher proportion of men dying.
+
 ## Weaknesses and next steps
-
 In exploring the implications of these findings, we gain valuable insights into the complex interplay of economic, social, and health factors that affect shelter residents.
@@
 
+\newpage
 \appendix
-# Appendix {-}
+# Appendix {#Appendix}
 ```{r echo=FALSE, message=FALSE, warning=FALSE}

### 2024-09-25 paper/paper.qmd
do style check add more llm usage change some content and typo add reference

--- 2024-09-25 paper/paper.qmd
@@
 date-format: long
-abstract: "Trough the analysis on the data about deaths of shelter residents, we found that the total number of deaths of shelter residences is increasing year by year.
+abstract: "Trough the analysis on the data about deaths of shelter residents, this paper found that the total number of deaths of shelter residences is increasing year by year.
 The proportion of male deaths is relatively higher.
-The findings provide insights into the factors affecting shelter residents' well-being and can guide policies to address homelessness, improve living conditions, and tackle health-related issues in this vulnerable population."
+And seasonal factors does not affect much on it.
+The findings provide insights into the factors affecting well-being of shelter residents and can guide policies to address homelessness, improve living conditions, and tackle health-related issues in this vulnerable population."
 format: pdf
@@
 # Read the analysis data of shelter residence death.
-shelter_death_data <- read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv", show_col_types = FALSE) 
+shelter_death_data <-
+  read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv",
+    show_col_types = FALSE
+  )
 ```
@@
 
-The rising number of deaths within Toronto's shelter system is a growing area of concern, reflecting broader societal challenges such as homelessness, mental health crises, and inadequate access to health care.
-Understanding the dynamics of these deaths is critical to developing effective policies and interventions aimed at reducing mortality in this vulnerable population.
-This paper focuses on analyzing the dataset related to deaths in Toronto's shelter residence, and seeks to identify the demographic trends and factors contributing to these unfortunate outcomes.
+The rising number of deaths in Toronto's shelter system is a growing area of concern, reflecting broader societal challenges including homelessness, mental health crises, and inadequate access to health care.
+Understanding the dynamics of these deaths is critical to developing effective policies and interventions aimed at reducing mortality in the vulnerable population.
+This paper focuses on analyzing the dataset related to deaths in Toronto's shelter residence, and seeks to identify the demographic trends and factors contributing to these outcomes.
 
-Despite the importance of this issue, there is a clear gap in the public's detailed understanding of who is being affected and how patterns of mortality are evolving over time.
-Due to privacy concerns, the dataset is limited in scope and lacks identifying information such as names, dates of birth, and specific causes of death.
-Also, the Non-binary genders are not included until 2020.
-Nevertheless, these aggregate data allow for an examination of key trends, such as the gender distribution of deaths and the change over time.
-These trends highlight critical disparities, particularly the higher proportion of male deaths and the steady increase in overall mortality number.
-
-This study aims to provide a visual and statistical analysis of the available data and to draw attention to the factors that may be contributing to the increased mortality.
-The analysis reveals two important findings: first, the number of deaths has steadily increased over the years, indicating a deteriorating situation within the shelter system; Second, men make up a larger proportion of the deceased, suggesting potential gender-specific vulnerabilities.
+Despite the limited scope of the dataset due to privacy concerns, which excludes identifiable information such as names, dates of birth, and specific causes of death, and the fact that non-binary genders were not included until 2020, the aggregated data allows for the examination of key trends.
+The paper finds that the total number of deaths from January 2007 to August 2024 showed an overall increasing trend, and that male deaths accounted for the largest share of all genders during this period.
+Besides, although the data shows that death occurs in winter most frequently, the result can be attributed to bias of the pandemic.
 
@@
 It contains key demographic variables and aggregates information about the number of deaths reported each year, categorized by gender.
-However, due to privacy concerns and respect for the confidentiality of individuals who have passed, the dataset excludes personal identifying information including names, birthdates, or specific locations of death.
-This limitation necessitated the use of data for analysis focusing on factors including gender distribution and time rather than locations and age.
+However, due to privacy concerns and respect for the confidentiality of individuals who have passed, the dataset excludes personal identifying information including names, age, or specific locations of death.
+This limitation necessitated the use of data for analysis focusing on factors including gender distribution and time.
+There are 212 observations in the original dataset in total and there is no special cleaning methods needed as the data is already clear enough.
+See Appendix @sec-appendix for the summary of the dataset.
 
@@
 Shelter operators are required to notify the City of Toronto within 24 hours of the death of a current or recent shelter resident and then submit a detailed written report within 30 days.
-This systematic approach, in place since 2007 and reviewed annually by Toronto Housing and Support Services, ensures timely and consistent data collection.
+(@toronto_deaths_shelter_residents) This systematic approach, in place since 2007 and reviewed annually by Toronto Housing and Support Services, ensures timely and consistent data collection.
 The structured reporting requirement imposes a level of discipline on shelter operators to ensure that deaths are not only reported in a timely manner, but also fully documented, thereby enhancing the integrity and credibility of the dataset.
@@
 However, certain limitations affect the completeness and reliability of the data.
-For example, due to privacy concerns, personal information such as names, dates of birth, and specific places of death are not collected.
-This limits the granularity of the data and can hinder deeper analysis, such as correlations with demographic or location-specific factors.
+For example, due to privacy concerns, personal information such as names, dates of birth, and specific places of death are not collected.(@toronto_deaths_shelter_residents) This limits the granularity of the data and can hinder deeper analysis, such as correlations with demographic or location-specific factors.
 In addition, shelter operators often do not have access to death certificates or information on causes of death, so the dataset lacks this critical context that could be important for understanding the factors contributing to mortality among shelter residents.
@@
 Improvements to the measurement process could include better integration with public health and legal entities to access more detailed information about deaths, including cause of death, while still protecting individual privacy.
-This improvement would provide a more holistic view of the factors influencing mortality rates among shelter residents, allowing for more informed policy decisions to address health and safety issues in shelters.
+This improvement would provide a more comprehensive view of the factors influencing mortality rates among shelter residents, allowing for more informed policy decisions to address health and safety issues in shelters.
 
@@
 
-Figure @fig-shelter-death-line, created by @citeR, shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year.
+@fig-shelter-death-line, created by @citeR, shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year.
 This layout allows readers to clearly compare the trend of deaths from one year to the next and observe potential seasonal changes.
@@
   geom_point(size = 2, color = "pink") + # add points on value in each year
-  labs(title = "Deaths in Toronto Shelters Over Time (2007.01-2024.08)",
+  labs(
+    title = "Deaths in Toronto Shelters Over Time (2007.01-2024.08)",
     x = "Month",
-       y = "Total Deaths") +  # add title and name of the axis
+    y = "Total Deaths"
+  ) + # add title and name of the axis
   theme_minimal() +
@@
     axis.title = element_text(size = 12),
-    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate horizontal axis labels to avoid overlapping
+    axis.text.x = element_text(angle = 45, hjust = 1)
+    # Rotate horizontal axis labels to avoid overlapping
   ) +
-  facet_wrap(~ year, ncol = 3)
-
+  facet_wrap(~year, ncol = 3)
 ```
@@
 ### Annual Deaths with Gender Composition
-Figure @fig-annual-death, created by @citeR, below illustrates the annual number of deaths from 2007 to 2023.
-And Figure @fig-shelter-death-bar, created by @citeR, illustrates the annual number of deaths of the same time period, with distinct color coding to represent gender distribution: pink for male, black for female, and gray for other gender identities, including transgender and non-binary.
+@fig-annual-death, created by @citeR, below illustrates the annual number of deaths from 2007 to 2023.
+And @fig-shelter-death-bar, created by @citeR, illustrates the annual number of deaths of the same time period, with distinct color coding to represent gender distribution: pink for male, black for female, and gray for other gender identities, including transgender and non-binary.
 
@@
 In addition, there is a noticeable upward trend in the overall number of deaths, suggesting a growing concern for the well-being of this vulnerable population.
-It can also be seen clearly from the line graph of Figure @fig-annual-death.
+It can also be seen clearly from the line graph of @fig-annual-death.
 Notably, 2021 stands out as the peak year for mortality.
-This can be seen clearly in Figure @fig-annual-death too.
+This can be seen clearly in @fig-annual-death too.
 This raises questions about the factors contributing to this spike and the broader implications for public health and social support systems.
@@
   ) %>%
-  pivot_longer(cols = c(male_deaths, female_deaths, other_deaths), 
+  pivot_longer(
+    cols = c(male_deaths, female_deaths, other_deaths),
     names_to = "gender",
-               values_to = "count")
+    values_to = "count"
+  )
 
-# Create bar chart showing annual death counts in Toronto shelters from 2007 to 2023.
-ggplot(annual_death_counts_bar, aes(x = factor(year), y = count, fill = gender)) +
+# Create bar chart showing annual death counts in Toronto shelters
+# from 2007 to 2023.
+ggplot(annual_death_counts_bar, aes(
+  x = factor(year),
+  y = count,
+  fill = gender
+)) +
   geom_bar(stat = "identity", position = "stack") +
-  labs(title = "Annual Death Counts in Toronto Shelters With Gender Distribution (2007-2023)",
+  labs(
+    title = "Annual Death Counts in Toronto
+       Shelters With Gender Distribution (2007-2023)",
     x = "year",
     y = "num of death",
-       fill = "gender") +
-  scale_fill_manual(values = c("male_deaths" = "pink", "female_deaths" = "black", "other_deaths" = "gray"), labels = c("Female", "male", "Other")) +  # change color and the text on the side
+    fill = "gender"
+  ) +
+  scale_fill_manual(
+    values = c(
+      "male_deaths" = "pink",
+      "female_deaths" = "black",
+      "other_deaths" = "gray"
+    ),
+    labels = c("Female", "male", "Other")
+  ) +
+  # change color and the text on the side
   theme_minimal() +
@@
     axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1)  # Rotate horizontal axis labels to avoid overlapping
+    axis.text.x = element_text(angle = 45, hjust = 1)
+    # Rotate horizontal axis labels to avoid overlapping
   )
@@
 ```{r fig-annual-death, fig.cap="Annual Deaths in Toronto Shelters (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
-
 # calculate the annual death
@@
   geom_point(size = 3, color = "pink") + # add points on value in each year
-  labs(title = "Annual Deaths in Toronto Shelters (2007-2023)",
+  labs(
+    title = "Annual Deaths in Toronto Shelters (2007-2023)",
     x = "Year",
-       y = "Total Deaths") +  # add title and name of the axis
+    y = "Total Deaths"
+  ) + # add title and name of the axis
   theme_minimal() +
@@
 ### Compariasion Over Month
-Figure @fig-monthly-death shows the aggregated monthly death counts for shelter residents from 2007 to 2023.
+@fig-monthly-death shows the aggregated monthly death counts for shelter residents from 2007 to 2023.
 This data was derived by summing the death counts for each January from 2007 to 2023, resulting in a comprehensive data set for January, and repeating this process for each of the other months, resulting in data for all twelve months.
@@
 ```{r fig-monthly-death, fig.cap="Monthly Mortality Trends Among Shelter Residents (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
-
 # calculate the annual death
@@
   geom_point(size = 3, color = "pink") + # add points on value in each year
-  labs(title = "Monthly Mortality Trends Among Shelter Residents (2007-2023)",
+  labs(
+    title = "Monthly Mortality Trends Among Shelter Residents (2007-2023)",
     x = "Month",
-       y = "Total Deaths") +  # as title and name of the axis
+    y = "Total Deaths"
+  ) + # as title and name of the axis
   theme_minimal() +
@@
 # Results{#sec-result}
-As time progresses, it is evident from Figures @fig-shelter-death-bar and @fig-annual-death that the number of deaths among shelter residents has increased.
+As time progresses, it is evident from @fig-shelter-death-bar and @fig-annual-death that the number of deaths among shelter residents has increased.
 Potential reasons for this trend may include changes in the economic landscape, shifts in the overall population, policy modifications, and factors related to the pandemic.
 
-Regarding monthly variations, while Figure @fig-monthly-death illustrates a noticeable peak in January and a low in September, a comprehensive view of the data in Figure @fig-shelter-death-line suggests that seasonal changes do not appear to have a significant correlation with the number of deaths.
+Regarding monthly variations, while @fig-monthly-death illustrates a noticeable peak in January and a low in September, a comprehensive view of the data in @fig-shelter-death-line suggests that seasonal changes do not appear to have a significant correlation with the number of deaths.
 One possible explanation for this phenomenon could be related to the timing of the pandemic's outbreak.
@@
 Our analysis shows that there was an upward trend in deaths during this period, with a higher proportion of men dying.
-## Weaknesses and next steps
 
 In exploring the implications of these findings, we gain valuable insights into the complex interplay of economic, social, and health factors that affect shelter residents.
-The data suggest that economic downturns, policy changes, and global health events may exacerbate the vulnerability of this population, warranting a reassessment of current support systems and interventions.
+The data suggest that economic downturns, policy changes, and global health events may exacerbate the vulnerability of this population.
 In addition, the lack of a clear seasonal pattern in mortality suggests that other factors, such as the pandemic, may be masking typical seasonal effects and warrant further investigation.
 
+## Weaknesses and next steps
 While the results highlight important issues, certain limitations must be acknowledged.
-The dataset used for this analysis may not fully capture the demographics of the shelter population, such as sex composition and health status.
+The dataset used for this analysis may not fully capture the demographics of the shelter population, such as its gender composition and health status.
 In addition, the analysis relied on reported deaths, which may be inaccurate or underreported.
@@
 
+\newpage
 \appendix
-# Appendix {-}
+# Appendix {#sec-appendix}
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
@@
 example_table %>%
-  kbl(caption = "Sample Table of Deaths in Toronto Shelters (2007-2023)",
-      col.names = c("ID", "Year", "Month", "Total decedents", "male", "female", "Others")) %>%
-  kable_styling(full_width = FALSE, position = "center", latex_options = "scale_down") %>%
+  kbl(
+    caption = "Sample Table of Deaths in Toronto Shelters (2007-2023)",
+    col.names = c(
+      "ID", "Year", "Month", "Total decedents",
+      "male", "female", "Others"
+    )
+  ) %>%
+  kable_styling(
+    full_width = FALSE,
+    position = "center",
+    latex_options = "scale_down"
+  ) %>%
   kable_classic(full_width = FALSE, html_font = "Cambria")
-
 ```
@@
     total_deaths = sum(total_decedents), # total death num
-    mean_deaths = round(mean(total_decedents), 2),  # average death num each month, rounded to 2 decimal places
-    median_deaths = round(median(total_decedents), 2),  # median, rounded
+    mean_deaths = round(mean(total_decedents), 2),
+    # average death num each month, rounded to 2 decimal places
+    median_deaths = round(median(total_decedents), 2),
+    # median, rounded
     min_deaths = min(total_decedents), # minimum value
     max_deaths = max(total_decedents), # maximum value
-    male_ratio = round(sum(male) / total_deaths * 100, 2),  # ratio of male death, rounded
-    female_ratio = round(sum(female) / total_deaths * 100, 2),  # ratio of female death, rounded
-    # other_ratio = round(sum(transgender_non_binary_two_spirit, na.rm = TRUE) / total_deaths * 100, 2)  # ratio of death of other genders, rounded
-    other_ratio = ifelse(any(is.na(shelter_death_data$transgender_non_binary_two_spirit)), NA,round(sum(transgender_non_binary_two_spirit, na.rm = TRUE) / total_deaths * 100, 2))  
+    male_ratio = round(sum(male) / total_deaths * 100, 2),
+    # ratio of male death, rounded
+    female_ratio = round(sum(female) / total_deaths * 100, 2),
+    # ratio of female death, rounded
+    other_ratio =
+      ifelse(any(is.na(shelter_death_data$transgender_non_binary_two_spirit)),
+        NA,
+        round(sum(transgender_non_binary_two_spirit,
+          na.rm = TRUE
+        ) / total_deaths * 100, 2)
+      )
     # ratio of death of other genders, rounded
@@
 summary_table %>%
-  kbl(caption = "Annual Summary of Deaths in Toronto Shelters (2007-2023)", 
-      col.names = c("Year", "Total Deaths", "Mean", "Median", "Min", "Max", "Male Ratio (%)", "Female Ratio (%)", "Other Ratio (%)")) %>%
-  kable_styling(full_width = FALSE, position = "center", latex_options = "scale_down") %>%
+  kbl(
+    caption = "Annual Summary of Deaths in Toronto Shelters (2007-2023)",
+    col.names = c(
+      "Year", "Total Deaths", "Mean", "Median", "Min", "Max",
+      "Male Ratio (%)", "Female Ratio (%)", "Other Ratio (%)"
+    )
+  ) %>%
+  kable_styling(
+    full_width = FALSE, position = "center",
+    latex_options = "scale_down"
+  ) %>%
   kable_classic(full_width = FALSE, html_font = "Cambria") %>%

### 2024-09-25 scripts/00-simulate_data.R

@@ -26,7 +26,7 @@ years <- format(dates, "%Y")
 
 months <- format(dates, "%b")
 
-simulated_shelter_death <- rpois(length(months),lambda =  3)
+simulated_shelter_death <- rpois(length(months), lambda = 3)
 
 male_death <- round(simulated_shelter_death * 0.65)
 
@@ -34,13 +34,17 @@ other_death <- round(simulated_shelter_death * 0.15)
 
 female_death <- simulated_shelter_death - male_death - other_death
 
-simulated_data <- data.frame(year = years, 
-                             month = months,
-                             total_death_num = simulated_shelter_death,
-                             male_death_num = male_death,
-                             female_death_num = female_death,
-                             other_death_num = other_death
+simulated_data <- data.frame(
+  year = years,
+  month = months,
+  total_death_num = simulated_shelter_death,
+  male_death_num = male_death,
+  female_death_num = female_death,
+  other_death_num = other_death
 )
 
 #### Write_csv
-write.csv(simulated_data, "data/simulated_data/simulated_shelter_residences_death_data.csv", row.names = FALSE)
+write.csv(simulated_data,
+  "data/simulated_data/simulated_shelter_residences_death_data.csv",
+  row.names = FALSE
+)

### 2024-09-25 scripts/01-download_data.R

@@ -21,12 +21,16 @@ package
 # get all resources for this package
 resources <- list_package_resources("deaths-of-shelter-residents")
 
-# identify datastore resources; by default, Toronto Open Data sets datastore resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(resources, tolower(format) %in% c('csv', 'geojson'))
+# identify datastore resources; by default, Toronto Open Data sets datastore
+# resource format to CSV for non-geospatial and GeoJSON for geospatial resources
+datastore_resources <- filter(
+  resources,
+  tolower(format) %in% c("csv", "geojson")
+)
 
 # load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number()==1) %>% get_resource()
+data <- filter(datastore_resources, row_number() == 1) %>% get_resource()
+
 #### Save data ####
 # change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(data, "data/raw_data/shelter_residences_death_raw_data.csv") 
-
+write_csv(data, "data/raw_data/shelter_residences_death_raw_data.csv")

### 2024-09-25 scripts/02-data_cleaning.R

@@ -11,7 +11,7 @@
 library(tidyverse)
 
 #### Get data ####
-raw_data <- read_csv("data/raw_data/shelter_residences_death_raw_data.csv") 
+raw_data <- read_csv("data/raw_data/shelter_residences_death_raw_data.csv")
 
 #### Clean data ####
 cleaned_data <-
@@ -21,9 +21,17 @@ cleaned_data <-
 
 
 cleaned_data <- cleaned_data %>%
-  mutate(transgender_non_binary_two_spirit = ifelse(transgender_non_binary_two_spirit == "n/a", NA, transgender_non_binary_two_spirit))
+  mutate(
+    transgender_non_binary_two_spirit =
+      ifelse(transgender_non_binary_two_spirit == "n/a",
+        NA,
+        transgender_non_binary_two_spirit
+      )
+  )
 
 
 #### Save data ####
-write_csv(cleaned_data, "data/analysis_data/shelter_residences_death_analysis_data.csv") 
-
+write_csv(
+  cleaned_data,
+  "data/analysis_data/shelter_residences_death_analysis_data.csv"
+)

### 2024-09-25 scripts/03-test_data.R

@@ -12,18 +12,25 @@
 library(tidyverse)
 
 #### Test data ####
-shelter_residences_death_data <- read_csv("data/simulated_data/simulated_shelter_residences_death_data.csv")
+shelter_residences_death_data <-
+  read_csv("data/simulated_data/simulated_shelter_residences_death_data.csv")
 
 # Test for negative numbers
 shelter_residences_death_data$total_death_num |> min() < 0
 
 # Test if month are valid
-valid_months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
+valid_months <- c(
+  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
+  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
+)
 all_correct <- all(months %in% valid_months)
 all_correct
 
 # Test if the total number of death matches the gender distribution
-test_total_num <- shelter_residences_death_data$total_death_num == shelter_residences_death_data$female_death_num + 
-  shelter_residences_death_data$male_death_num + shelter_residences_death_data$other_death_num
+test_total_num <-
+  shelter_residences_death_data$total_death_num ==
+    shelter_residences_death_data$female_death_num +
+      shelter_residences_death_data$male_death_num +
+      shelter_residences_death_data$other_death_num
 all_match <- all(test_total_num)
 all_match

### 2024-09-25 paper/paper.qmd
add more cross-reference chang part of the paper text content

@@
 
-**Data Quality and Cleaning**: The dataset required careful cleaning to address issues such as missing values and inconsistencies, particularly in the reporting of deaths by gender.
-Although high-level cleaning was conducted to ensure the reliability of the data, the absence of certain variables may still introduce biases or limit the interpretability of the results.
+**Data Quality and Cleaning**: The dataset misses values of transgender/Non-binary/two-spirit decedents until 2020 due to lack of awareness of gender identification.
+The missing values only affect the research in gender vulnerability field and it is hard to be fulfilled.
+In our research, the missing value is replaced with 0 mostly.
+The value keeps being NA when calculate the death ratio of each gender.
 
@@
 
-### Trend Over Time
-
+### Trend Over Time {#sec-time}
 @fig-shelter-death-line, created by @citeR, shows the total number of deaths of Toronto shelter residents each year from January 2007 to August 2024, divided into 15 sub-graphs, arranged in 5 rows and 3 columns by year.
@@
 The purpose of this visualization is to analyze whether certain months have higher mortality rates.
-It opens the door to exploring potential seasonal influences on mortality rates, such as the impact of extreme temperatures or adverse weather conditions, which could contribute to increased mortality among shelter residents.
+It helps to exploring potential seasonal influences on mortality rates, such as the impact of extreme temperatures or adverse weather conditions, which could contribute to increased mortality among shelter residents.
 
-The current findings, shown in the line graph, indicate that January and November have the highest mortality rates, while March and September have the lowest mortality rates.
+The current findings, shown in the line graph, indicate that January and November have the highest mortality rates, while March and September have the lowest mortality rates, which was different in what we find in @sec-time.
 These findings highlight the need for further research into the factors influencing these monthly variations, which could inform targeted interventions to improve the health and safety of shelter residents.

### 2024-09-25 paper/paper.qmd
add more llm usage

@@
 The current findings, shown in the line graph, indicate that January and November have the highest mortality rates, while March and September have the lowest mortality rates, which was different in what we find in @sec-time.
-These findings highlight the need for further research into the factors influencing these monthly variations, which could inform targeted interventions to improve the health and safety of shelter residents.
+These insights highlight the need for further investigation into the factors influencing these monthly variations to inform targeted interventions and improve the health and safety of shelter residents.
 

### 2024-09-25 paper/paper.qmd
formatting add llm usage

@@
 \appendix
+
 # Appendix {#sec-appendix}
+Table 1 shows the a sample how data is like in the original dataset from @toronto_shelter_deaths provided by @opendatatoronto. It is the first 5 lines in the dataset.
+
+Table 2 shows the summary of our dataset.
+It contains the total decedents number, monthly average deaths, minimum, maximum and median of the each year.
+It also provide the ratio of different gender in the shelter decedents each year.
+
 ```{r echo=FALSE, message=FALSE, warning=FALSE}
-# Get first 10 lines in the data set as a sample
+# Get first 20 lines in the data set as a sample
 example_table <- shelter_death_data %>%
-  slice(1:10)
+  slice(1:5)
 
@@
 \newpage
-
 # References

### 2024-09-25 paper/paper.qmd
add intext reference in paper

@@
 ### Compariasion Over Month
-@fig-monthly-death shows the aggregated monthly death counts for shelter residents from 2007 to 2023.
+@fig-monthly-death, created by @citeR, shows the aggregated monthly death counts for shelter residents from 2007 to 2023.
 This data was derived by summing the death counts for each January from 2007 to 2023, resulting in a comprehensive data set for January, and repeating this process for each of the other months, resulting in data for all twelve months.
@@
 # Appendix {#sec-appendix}
-Table 1 shows the a sample how data is like in the original dataset from @toronto_shelter_deaths provided by @opendatatoronto. It is the first 5 lines in the dataset.
+Table 1, created by @citeR, shows the a sample how data is like in the original dataset from @toronto_shelter_deaths provided by @opendatatoronto. It is the first 5 lines in the dataset.
 
-Table 2 shows the summary of our dataset.
+Table 2, created by @citeR, shows the summary of our dataset.
 It contains the total decedents number, monthly average deaths, minimum, maximum and median of the each year.

### 2025-06-20 scripts/00-simulate_data.R
scripts to python

@@ -1,50 +1,0 @@
-#### Preamble ####
-# Purpose: Simulates data
-# Author: Xinqi Yue
-# Date: 23 September 2024
-# Contact: xinqi.yue@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: None
-# Any other information needed? None
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(dplyr)
-library(lubridate)
-
-#### Simulate data ####
-set.seed(507)
-
-# Simulate data
-dates <- seq.Date(
-  from = as.Date("2007-01-01"),
-  to = as.Date("2024-08-01"),
-  by = "month"
-)
-years <- format(dates, "%Y")
-
-months <- format(dates, "%b")
-
-simulated_shelter_death <- rpois(length(months), lambda = 3)
-
-male_death <- round(simulated_shelter_death * 0.65)
-
-other_death <- round(simulated_shelter_death * 0.15)
-
-female_death <- simulated_shelter_death - male_death - other_death
-
-simulated_data <- data.frame(
-  year = years,
-  month = months,
-  total_death_num = simulated_shelter_death,
-  male_death_num = male_death,
-  female_death_num = female_death,
-  other_death_num = other_death
-)
-
-#### Write_csv
-write.csv(simulated_data,
-  "data/simulated_data/simulated_shelter_residences_death_data.csv",
-  row.names = FALSE
-)

### 2025-06-20 scripts/00-simulate_data.py

@@ -1,0 +1,44 @@
+#### Preamble ####
+# Purpose: Simulates data
+# Author: Xinqi Yue
+# Date: 23 September 2024
+# Contact: xinqi.yue@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+# Any other information needed? None
+
+#### Workspace setup ####
+import pandas as pd
+import numpy as np
+
+#### Simulate data ####
+np.random.seed(507)
+
+# Generate monthly dates from Jan 2007 to Aug 2024
+dates = pd.date_range(start="2007-01-01", end="2024-08-01", freq='MS')
+years = dates.year
+months = dates.strftime("%b")
+
+# Simulate deaths using Poisson distribution
+simulated_shelter_death = np.random.poisson(lam=3, size=len(dates))
+
+# Allocate gender breakdown
+male_death = np.round(simulated_shelter_death * 0.65).astype(int)
+other_death = np.round(simulated_shelter_death * 0.15).astype(int)
+female_death = simulated_shelter_death - male_death - other_death
+
+# Create DataFrame
+simulated_data = pd.DataFrame({
+    "year": years,
+    "month": months,
+    "total_death_num": simulated_shelter_death,
+    "male_death_num": male_death,
+    "female_death_num": female_death,
+    "other_death_num": other_death
+})
+
+#### Write CSV ####
+simulated_data.to_csv(
+    "data/simulated_data/simulated_shelter_residences_death_data.csv",
+    index=False
+)

### 2025-06-20 scripts/01-download_data.R

@@ -1,36 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from Open Data Toronto
-# Author: Xinqi Yue
-# Date: 23 September 2024
-# Contact: xinqi.yue@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: None
-# Any other information needed? None
-
-
-#### Workspace setup ####
-library(opendatatoronto)
-library(tidyverse)
-library(dplyr)
-
-#### Download data ####
-# get package
-package <- show_package("deaths-of-shelter-residents")
-package
-
-# get all resources for this package
-resources <- list_package_resources("deaths-of-shelter-residents")
-
-# identify datastore resources; by default, Toronto Open Data sets datastore
-# resource format to CSV for non-geospatial and GeoJSON for geospatial resources
-datastore_resources <- filter(
-  resources,
-  tolower(format) %in% c("csv", "geojson")
-)
-
-# load the first datastore resource as a sample
-data <- filter(datastore_resources, row_number() == 1) %>% get_resource()
-
-#### Save data ####
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(data, "data/raw_data/shelter_residences_death_raw_data.csv")

### 2025-06-20 scripts/01-download_data.py

@@ -1,0 +1,33 @@
+#### Preamble ####
+# Purpose: Downloads and saves the data from Open Data Toronto
+# Author: Xinqi Yue
+# Date: 23 September 2024
+# Contact: xinqi.yue@mail.utoronto.ca
+# License: MIT
+# Pre-requisites: None
+# Any other information needed? None
+
+import requests
+import pandas as pd
+from io import StringIO
+	
+base_url = "https://ckan0.cf.opendata.inter.prod-toronto.ca"
+ 
+# Datasets are called "packages". Each package can contain many "resources"
+# To retrieve the metadata for this package and its resources, use the package name in this page's URL:
+url = base_url + "/api/3/action/package_show"
+params = { "id": "deaths-of-shelter-residents"}
+package = requests.get(url, params = params).json()
+ 
+# To get resource data:
+for idx, resource in enumerate(package["result"]["resources"]):
+ 
+       # for datastore_active resources:
+       if resource["datastore_active"]:
+ 
+           # To get all records in CSV format:
+           url = base_url + "/datastore/dump/" + resource["id"]
+           resource_dump_data = requests.get(url).text
+           data = pd.read_csv(StringIO(resource_dump_data))
+           data.to_csv("data/raw_data/shelter_residences_death_raw_data.csv", index=False)
+           break

### 2025-06-20 scripts/02-data_cleaning.R

@@ -1,37 +1,0 @@
-#### Preamble ####
-# Purpose: Cleans the raw marriage data into an analysis dataset
-# Author: Xinqi Yue
-# Date: 23 September 2024
-# Contact: xinqi.yue@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: None
-# Any other information needed? None
-
-#### Workspace setup ####
-library(tidyverse)
-
-#### Get data ####
-raw_data <- read_csv("data/raw_data/shelter_residences_death_raw_data.csv")
-
-#### Clean data ####
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  tidyr::drop_na()
-
-
-cleaned_data <- cleaned_data %>%
-  mutate(
-    transgender_non_binary_two_spirit =
-      ifelse(transgender_non_binary_two_spirit == "n/a",
-        NA,
-        transgender_non_binary_two_spirit
-      )
-  )
-
-
-#### Save data ####
-write_csv(
-  cleaned_data,
-  "data/analysis_data/shelter_residences_death_analysis_data.csv"
-)

### 2025-06-20 scripts/02-data_cleaning.py

@@ -1,0 +1,35 @@
+#### Preamble ####
+# Purpose: Cleans the raw marriage data into an analysis dataset
+# Author: Xinqi Yue
+# Date: 23 September 2024
+# Contact: xinqi.yue@mail.utoronto.ca
+# License: MIT
+
+#### Workspace setup ####
+import pandas as pd
+
+#### Get data ####
+raw_data = pd.read_csv("data/raw_data/shelter_residences_death_raw_data.csv")
+
+#### Clean column names manually (like janitor::clean_names) ####
+raw_data.columns = (
+    raw_data.columns
+    .str.strip()
+    .str.lower()
+    .str.replace(' ', '_')
+    .str.replace(r'[^\w\s]', '', regex=True)
+)
+
+raw_data["transgendernonbinarytwospirit"] = raw_data["transgendernonbinarytwospirit"].fillna("NA")
+
+#### Drop rows with missing values ####
+cleaned_data = raw_data.dropna()
+
+#to match original version
+cleaned_data = cleaned_data[cleaned_data["_id"] < 213]
+
+#### Save cleaned data ####
+cleaned_data.to_csv(
+    "data/analysis_data/shelter_residences_death_analysis_data.csv",
+    index=False
+)

### 2025-06-21 paper/paper.qmd
changed paper scripts

@@
 
-```{r}
+```{python}
 #| include: false
@@
 
-library(kableExtra)
-library(tidyverse)
-library(palmerpenguins)
+import pandas as pd
+import matplotlib.pyplot as plt
+import numpy as np
 
-# Read the analysis data of shelter residence death.
-shelter_death_data <-
-  read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv",
-    show_col_types = FALSE
-  )
+# Read the analysis data of shelter residence deaths
+shelter_death_data = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
 ```
@@
 This layout allows readers to clearly compare the trend of deaths from one year to the next and observe potential seasonal changes.
+```{python}
+#| label: fig-shelter-death-line
+#| fig-cap: "Deaths in Toronto Shelters Over Time (2007.01–2024.08)"
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-width: 9
+#| fig-height: 6
 
-```{r fig-shelter-death-line, fig.cap="Deaths in Toronto Shelters Over Time (2007.01-2024.08)", echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=6}
-# Sort the month data in order
-shelter_death_data$month <- factor(shelter_death_data$month, levels = month.abb)
+# Load the data
+df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
 
-# Get line graph, showing the death trend
-ggplot(shelter_death_data, aes(x = month, y = total_decedents, group = year)) +
-  geom_line(size = 1, color = "black") + # get line graph
-  geom_point(size = 2, color = "pink") + # add points on value in each year
-  labs(
-    title = "Deaths in Toronto Shelters Over Time (2007.01-2024.08)",
-    x = "Month",
-    y = "Total Deaths"
-  ) + # add title and name of the axis
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
-    axis.title = element_text(size = 12),
-    axis.text.x = element_text(angle = 45, hjust = 1)
-    # Rotate horizontal axis labels to avoid overlapping
-  ) +
-  facet_wrap(~year, ncol = 3)
+# Filter for years 2007–2023
+filtered = df[(df["year"] >= 2007) & (df["year"] <= 2023)].copy()
+
+# Group by year and compute summary stats
+summary = filtered.groupby("year").agg(
+    total_deaths=("total_decedents", "sum"),
+    mean_deaths=("total_decedents", lambda x: round(x.mean(), 2)),
+    median_deaths=("total_decedents", lambda x: round(x.median(), 2)),
+    min_deaths=("total_decedents", "min"),
+    max_deaths=("total_decedents", "max"),
+    male_sum=("male", "sum"),
+    female_sum=("female", "sum"),
+    other_sum=("transgendernonbinarytwospirit", lambda x: x.sum(skipna=True))
+).reset_index()
+
+# Calculate gender ratios
+summary["male_ratio"] = round(summary["male_sum"] / summary["total_deaths"] * 100, 2)
+summary["female_ratio"] = round(summary["female_sum"] / summary["total_deaths"] * 100, 2)
+
+# Handle possible NA case in other gender ratio
+if df["transgendernonbinarytwospirit"].isna().any():
+    summary["other_ratio"] = np.nan
+else:
+    summary["other_ratio"] = round(summary["other_sum"] / summary["total_deaths"] * 100, 2)
+
+# Keep only relevant columns and rename for display
+summary_table = summary[[
+    "year", "total_deaths", "mean_deaths", "median_deaths", "min_deaths", "max_deaths",
+    "male_ratio", "female_ratio", "other_ratio"
+]].rename(columns={
+    "year": "Year",
+    "total_deaths": "Total Deaths",
+    "mean_deaths": "Mean",
+    "median_deaths": "Median",
+    "min_deaths": "Min",
+    "max_deaths": "Max",
+    "male_ratio": "Male Ratio (%)",
+    "female_ratio": "Female Ratio (%)",
+    "other_ratio": "Other Ratio (%)"
+})
+
+# Display styled table with caption
+summary_table.style.set_caption("Annual Summary of Deaths in Toronto Shelters (2007–2023)") \
+    .set_table_styles([
+        {"selector": "caption", "props": [("caption-side", "top"), ("font-size", "110%"), ("font-weight", "bold")]},
+        {"selector": "th", "props": [("text-align", "center")]},
+        {"selector": "td", "props": [("text-align", "center")]}
+    ]) \
+    .set_properties(**{"font-family": "Cambria"}) \
+    .format(na_rep="NA")
 ```
@@
 
-```{r fig-shelter-death-bar, fig.cap="Annual Death Counts in Toronto Shelters With Gender Distribution (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=6}
-annual_death_counts_bar <- shelter_death_data %>%
-  filter(year >= 2007 & year <= 2023) %>%
-  group_by(year) %>%
-  summarise(
-    total_deaths = sum(total_decedents),
-    male_deaths = sum(male),
-    female_deaths = sum(female),
-    other_deaths = sum(transgender_non_binary_two_spirit, na.rm = TRUE)
-  ) %>%
-  pivot_longer(
-    cols = c(male_deaths, female_deaths, other_deaths),
-    names_to = "gender",
-    values_to = "count"
+```{python} 
+#| label: fig-shelter-death-bar
+#| fig-cap: "Annual Death Counts in Toronto Shelters With Gender Distribution (2007–2023)"
+#| echo: false
+#| message: false
+#| warning: false
+#| fig-width: 10
+#| fig-height: 6
+
+import pandas as pd
+import matplotlib.pyplot as plt
+
+# Load the data
+df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
+
+# Filter for years 2007–2023
+df = df[(df["year"] >= 2007) & (df["year"] <= 2023)]
+
+# Group and sum by year
+grouped = df.groupby("year").agg(
+    male_deaths=("male", "sum"),
+    female_deaths=("female", "sum"),
+    other_deaths=("transgendernonbinarytwospirit", "sum")
+).fillna(0)
+
+# Convert to long format
+long_df = grouped.reset_index().melt(id_vars="year", var_name="gender", value_name="count")
+
+# Set color mapping and label override
+color_map = {
+    "male_deaths": "pink",
+    "female_deaths": "black",
+    "other_deaths": "gray"
+}
+label_map = {
+    "male_deaths": "male",
+    "female_deaths": "Female",
+    "other_deaths": "Other"
+}
+
+# Pivot for stacked bar
+pivot_df = long_df.pivot(index="year", columns="gender", values="count").fillna(0)
+pivot_df = pivot_df[color_map.keys()]  # Ensure column order
+
+# Plot
+fig, ax = plt.subplots(figsize=(10, 6))
+
+bottom = None
+for gender in pivot_df.columns:
+    ax.bar(pivot_df.index, pivot_df[gender], label=label_map[gender],
+           bottom=bottom, color=color_map[gender])
+    bottom = pivot_df[gender] if bottom is None else bottom + pivot_df[gender]
+
+# Labels and title
+ax.set_title("Annual Death Counts in Toronto Shelters With Gender Distribution (2007–2023)",
+             fontsize=16, fontweight='bold', ha='center')
+ax.set_xlabel("Year", fontsize=12)
+ax.set_ylabel("Num of Death", fontsize=12)
+ax.set_xticks(pivot_df.index)
+ax.set_xticklabels(pivot_df.index, rotation=45)
+ax.tick_params(axis='both', labelsize=10)
+
+# Legend and theme
+ax.legend(title="Gender", title_fontsize=11)
+ax.grid(True, axis='y', linestyle='--', alpha=0.7)
+ax.set_axisbelow(True)
+
+plt.tight_layout()
+plt.show()
+
+```
+
+```{python}
+#| label: fig-annual-death
+#| fig-cap: "Annual Deaths in Toronto Shelters (2007–2023)"
+#| echo: false
+#| message: false
+#| warning: false
+
+import pandas as pd
+import matplotlib.pyplot as plt
+
+# Load the data
+df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
+
+# Filter years and calculate total annual deaths
+annual_deaths = (
+    df[(df["year"] >= 2007) & (df["year"] <= 2023)]
+    .groupby("year", as_index=False)
+    .agg(total_yearly_deaths=("total_decedents", "sum"))
 )
 
-# Create bar chart showing annual death counts in Toronto shelters
-# from 2007 to 2023.
-ggplot(annual_death_counts_bar, aes(
-  x = factor(year),
-  y = count,
-  fill = gender
-)) +
-  geom_bar(stat = "identity", position = "stack") +
-  labs(
-    title = "Annual Death Counts in Toronto
-       Shelters With Gender Distribution (2007-2023)",
-    x = "year",
-    y = "num of death",
-    fill = "gender"
-  ) +
-  scale_fill_manual(
-    values = c(
-      "male_deaths" = "pink",
-      "female_deaths" = "black",
-      "other_deaths" = "gray"
-    ),
-    labels = c("Female", "male", "Other")
-  ) +
-  # change color and the text on the side
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
-    axis.title = element_text(size = 12),
-    axis.text = element_text(size = 10),
-    axis.text.x = element_text(angle = 45, hjust = 1)
-    # Rotate horizontal axis labels to avoid overlapping
-  )
-```
+# Plot line chart
+fig, ax = plt.subplots(figsize=(7, 4))
 
-```{r fig-annual-death, fig.cap="Annual Deaths in Toronto Shelters (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
-# calculate the annual death
-annual_deaths <- shelter_death_data %>%
-  filter(year >= 2007 & year <= 2023) %>%
-  group_by(year) %>%
-  summarise(total_yearly_deaths = sum(total_decedents), na.rm = TRUE)
+ax.plot(annual_deaths["year"], annual_deaths["total_yearly_deaths"],
+        color="black", linewidth=1)
+ax.scatter(annual_deaths["year"], annual_deaths["total_yearly_deaths"],
+           color="pink", s=40)
 
-# get line graph, showing the annual death trend
-ggplot(annual_deaths, aes(x = year, y = total_yearly_deaths)) +
-  geom_line(size = 1, color = "black") + # get line graph
-  geom_point(size = 3, color = "pink") + # add points on value in each year
-  labs(
-    title = "Annual Deaths in Toronto Shelters (2007-2023)",
-    x = "Year",
-    y = "Total Deaths"
-  ) + # add title and name of the axis
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
-    axis.title = element_text(size = 10),
-    axis.text = element_text(size = 8)
-  )
+# Labels and title
+ax.set_title("Annual Deaths in Toronto Shelters (2007–2023)",
+             fontsize=12, fontweight='bold', ha='center')
+ax.set_xlabel("Year", fontsize=10)
+ax.set_ylabel("Total Deaths", fontsize=10)
+ax.tick_params(axis='both', labelsize=8)
+
+# Apply minimal-style grid
+ax.grid(True, linestyle='--', linewidth=0.5, alpha=0.7)
+ax.set_axisbelow(True)
+
+plt.tight_layout()
+plt.show()
+
 ```
@@
 
-```{r fig-monthly-death, fig.cap="Monthly Mortality Trends Among Shelter Residents (2007-2023)", echo=FALSE, message=FALSE, warning=FALSE}
-# calculate the annual death
-month_deaths <- shelter_death_data %>%
-  filter(year >= 2007 & year <= 2023) %>%
-  group_by(month) %>%
-  summarise(total_monthly_deaths = sum(total_decedents))
+```{python}
+#| label: fig-monthly-death
+#| fig-cap: "Monthly Mortality Trends Among Shelter Residents (2007–2023)"
+#| echo: false
+#| message: false
+#| warning: false
 
-month_deaths$month <- factor(month_deaths$month, levels = month.abb)
+# Load the data
+df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
 
-# get line graph, showing the annual death trend
-ggplot(month_deaths, aes(x = month, y = total_monthly_deaths, group = 1)) +
-  geom_line(size = 1, color = "black") + # get line graph
-  geom_point(size = 3, color = "pink") + # add points on value in each year
-  labs(
-    title = "Monthly Mortality Trends Among Shelter Residents (2007-2023)",
-    x = "Month",
-    y = "Total Deaths"
-  ) + # as title and name of the axis
-  theme_minimal() +
-  theme(
-    plot.title = element_text(hjust = 0.5, size = 12, face = "bold"),
-    axis.title = element_text(size = 10),
-    axis.text = element_text(size = 8)
+# Filter for relevant years and group by month
+month_deaths = (
+    df[(df["year"] >= 2007) & (df["year"] <= 2023)]
+    .groupby("month", as_index=False)
+    .agg(total_monthly_deaths=("total_decedents", "sum"))
 )
+
+# Ensure correct month order
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+month_deaths["month"] = pd.Categorical(month_deaths["month"],
+                                       categories=month_order,
+                                       ordered=True)
+month_deaths = month_deaths.sort_values("month")
+
+# Plot
+fig, ax = plt.subplots(figsize=(7, 4))
+
+ax.plot(month_deaths["month"], month_deaths["total_monthly_deaths"],
+        color="black", linewidth=1)
+ax.scatter(month_deaths["month"], month_deaths["total_monthly_deaths"],
+           color="pink", s=40)
+
+# Labels and theme
+ax.set_title("Monthly Mortality Trends Among Shelter Residents (2007–2023)",
+             fontsize=12, fontweight='bold', ha='center')
+ax.set_xlabel("Month", fontsize=10)
+ax.set_ylabel("Total Deaths", fontsize=10)
+ax.tick_params(axis='both', labelsize=8)
+
+ax.grid(True, linestyle='--', linewidth=0.5, alpha=0.7)
+ax.set_axisbelow(True)
+
+plt.tight_layout()
+plt.show()
+
 ```
@@
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
-# Get first 20 lines in the data set as a sample
-example_table <- shelter_death_data %>%
-  slice(1:5)
+```{python}
+#| echo: false
+#| message: false
+#| warning: false
 
-example_table %>%
-  kbl(
-    caption = "Sample Table of Deaths in Toronto Shelters (2007-2023)",
-    col.names = c(
+# Load data
+df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
+
+# Take first 5 rows as sample
+example_table = df.iloc[:5].copy()
+
+# Rename columns for display
+example_table.columns = [
     "ID", "Year", "Month", "Total decedents",
     "male", "female", "Others"
-    )
-  ) %>%
-  kable_styling(
-    full_width = FALSE,
-    position = "center",
-    latex_options = "scale_down"
-  ) %>%
-  kable_classic(full_width = FALSE, html_font = "Cambria")
+] + list(example_table.columns[7:])  # In case there are more columns
+
+# Display the table (nicely rendered in Quarto)
+example_table.style.set_caption("Sample Table of Deaths in Toronto Shelters (2007–2023)") \
+    .set_table_styles(
+        [{"selector": "caption", "props": [("caption-side", "top"), ("font-size", "110%"), ("font-weight", "bold")]}]
+    ) \
+    .set_properties(**{"text-align": "center", "font-family": "Cambria"})
+
 ```
 
-```{r echo=FALSE, message=FALSE, warning=FALSE}
-# summarize the data group by year
-summary_table <- shelter_death_data %>%
-  filter(year >= 2007 & year <= 2023) %>%
-  group_by(year) %>%
-  summarise(
-    total_deaths = sum(total_decedents), # total death num
-    mean_deaths = round(mean(total_decedents), 2),
-    # average death num each month, rounded to 2 decimal places
-    median_deaths = round(median(total_decedents), 2),
-    # median, rounded
-    min_deaths = min(total_decedents), # minimum value
-    max_deaths = max(total_decedents), # maximum value
-    male_ratio = round(sum(male) / total_deaths * 100, 2),
-    # ratio of male death, rounded
-    female_ratio = round(sum(female) / total_deaths * 100, 2),
-    # ratio of female death, rounded
-    other_ratio =
-      ifelse(any(is.na(shelter_death_data$transgender_non_binary_two_spirit)),
-        NA,
-        round(sum(transgender_non_binary_two_spirit,
-          na.rm = TRUE
-        ) / total_deaths * 100, 2)
-      )
-    # ratio of death of other genders, rounded
-  )
-summary_table %>%
-  kbl(
-    caption = "Annual Summary of Deaths in Toronto Shelters (2007-2023)",
-    col.names = c(
-      "Year", "Total Deaths", "Mean", "Median", "Min", "Max",
-      "Male Ratio (%)", "Female Ratio (%)", "Other Ratio (%)"
-    )
-  ) %>%
-  kable_styling(
-    full_width = FALSE, position = "center",
-    latex_options = "scale_down"
-  ) %>%
-  kable_classic(full_width = FALSE, html_font = "Cambria") %>%
-  add_header_above(c(" " = 1, "Deaths Summary" = 5, "Gender Ratios" = 3))
+```{python}
+#| echo: false
+#| message: false
+#| warning: false
+
+# Load the data
+df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
+
+# Filter for years 2007–2023
+filtered = df[(df["year"] >= 2007) & (df["year"] <= 2023)].copy()
+
+# Group by year and compute summary stats
+summary = filtered.groupby("year").agg(
+    total_deaths=("total_decedents", "sum"),
+    mean_deaths=("total_decedents", lambda x: round(x.mean(), 2)),
+    median_deaths=("total_decedents", lambda x: round(x.median(), 2)),
+    min_deaths=("total_decedents", "min"),
+    max_deaths=("total_decedents", "max"),
+    male_sum=("male", "sum"),
+    female_sum=("female", "sum"),
+    other_sum=("transgendernonbinarytwospirit", lambda x: x.sum(skipna=True))
+).reset_index()
+
+# Calculate gender ratios
+summary["male_ratio"] = round(summary["male_sum"] / summary["total_deaths"] * 100, 2)
+summary["female_ratio"] = round(summary["female_sum"] / summary["total_deaths"] * 100, 2)
+
+# Handle possible NA case in other gender ratio
+if df["transgendernonbinarytwospirit"].isna().any():
+    summary["other_ratio"] = np.nan
+else:
+    summary["other_ratio"] = round(summary["other_sum"] / summary["total_deaths"] * 100, 2)
+
+# Keep only relevant columns and rename for display
+summary_table = summary[[
+    "year", "total_deaths", "mean_deaths", "median_deaths", "min_deaths", "max_deaths",
+    "male_ratio", "female_ratio", "other_ratio"
+]].rename(columns={
+    "year": "Year",
+    "total_deaths": "Total Deaths",
+    "mean_deaths": "Mean",
+    "median_deaths": "Median",
+    "min_deaths": "Min",
+    "max_deaths": "Max",
+    "male_ratio": "Male Ratio (%)",
+    "female_ratio": "Female Ratio (%)",
+    "other_ratio": "Other Ratio (%)"
+})
+
+# Display styled table with caption
+summary_table.style.set_caption("Annual Summary of Deaths in Toronto Shelters (2007–2023)") \
+    .set_table_styles([
+        {"selector": "caption", "props": [("caption-side", "top"), ("font-size", "110%"), ("font-weight", "bold")]},
+        {"selector": "th", "props": [("text-align", "center")]},
+        {"selector": "td", "props": [("text-align", "center")]}
+    ]) \
+    .set_properties(**{"font-family": "Cambria"}) \
+    .format(na_rep="NA")
 ```

### 2025-06-21 scripts/03-test_data.py

@@ -5,32 +5,26 @@
 # Contact: xinqi.yue@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: Need to have simulated data
-# Any other information needed? None
-
 
 #### Workspace setup ####
-library(tidyverse)
+import pandas as pd
 
 #### Test data ####
-shelter_residences_death_data <-
-  read_csv("data/simulated_data/simulated_shelter_residences_death_data.csv")
+df = pd.read_csv("data/simulated_data/simulated_shelter_residences_death_data.csv")
 
-# Test for negative numbers
-shelter_residences_death_data$total_death_num |> min() < 0
+# Test for negative numbers in total deaths
+has_negative_total = (df["total_death_num"] < 0).any()
+print("Any negative total death count:", has_negative_total)
 
-# Test if month are valid
-valid_months <- c(
-  "Jan", "Feb", "Mar", "Apr", "May", "Jun",
-  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
-)
-all_correct <- all(months %in% valid_months)
-all_correct
+# Test if months are valid
+valid_months = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+all_months_valid = df["month"].isin(valid_months).all()
+print("All months valid:", all_months_valid)
 
-# Test if the total number of death matches the gender distribution
-test_total_num <-
-  shelter_residences_death_data$total_death_num ==
-    shelter_residences_death_data$female_death_num +
-      shelter_residences_death_data$male_death_num +
-      shelter_residences_death_data$other_death_num
-all_match <- all(test_total_num)
-all_match
+# Test if total_death_num equals sum of gender breakdowns
+gender_sum_matches = (df["total_death_num"] ==
+                      df["male_death_num"] +
+                      df["female_death_num"] +
+                      df["other_death_num"]).all()
+print("All gender sums match total:", gender_sum_matches)

### 2025-06-21 paper/paper.qmd
fixed issue w/ axes

@@
 # Load the data
-df = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
+shelter_death_data = pd.read_csv("../data/analysis_data/shelter_residences_death_analysis_data.csv")
 
-# Filter for years 2007–2023
-filtered = df[(df["year"] >= 2007) & (df["year"] <= 2023)].copy()
+# Order the months
+month_order = ["Jan", "Feb", "Mar", "Apr", "May", "Jun",
+               "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"]
+shelter_death_data["month"] = pd.Categorical(
+    shelter_death_data["month"], categories=month_order, ordered=True
+)
 
-# Group by year and compute summary stats
-summary = filtered.groupby("year").agg(
-    total_deaths=("total_decedents", "sum"),
-    mean_deaths=("total_decedents", lambda x: round(x.mean(), 2)),
-    median_deaths=("total_decedents", lambda x: round(x.median(), 2)),
-    min_deaths=("total_decedents", "min"),
-    max_deaths=("total_decedents", "max"),
-    male_sum=("male", "sum"),
-    female_sum=("female", "sum"),
-    other_sum=("transgendernonbinarytwospirit", lambda x: x.sum(skipna=True))
-).reset_index()
+# Prepare facet layout: 3 columns
+years = sorted(shelter_death_data["year"].unique())
+ncols = 3
+nrows = (len(years) + ncols - 1) // ncols
 
-# Calculate gender ratios
-summary["male_ratio"] = round(summary["male_sum"] / summary["total_deaths"] * 100, 2)
-summary["female_ratio"] = round(summary["female_sum"] / summary["total_deaths"] * 100, 2)
+# Set sharey=True to unify Y-axis scale across all facets
+fig, axes = plt.subplots(nrows=nrows, ncols=ncols, figsize=(9, nrows * 2.5), sharey=True)
+axes = axes.flatten()
 
-# Handle possible NA case in other gender ratio
-if df["transgendernonbinarytwospirit"].isna().any():
-    summary["other_ratio"] = np.nan
-else:
-    summary["other_ratio"] = round(summary["other_sum"] / summary["total_deaths"] * 100, 2)
+# Plot each year
+for i, year in enumerate(years):
+    ax = axes[i]
+    yearly_data = shelter_death_data[shelter_death_data["year"] == year].sort_values("month")
+    ax.plot(yearly_data["month"], yearly_data["total_decedents"], color="black", linewidth=1)
+    ax.scatter(yearly_data["month"], yearly_data["total_decedents"], color="pink", s=40)
 
-# Keep only relevant columns and rename for display
-summary_table = summary[[
-    "year", "total_deaths", "mean_deaths", "median_deaths", "min_deaths", "max_deaths",
-    "male_ratio", "female_ratio", "other_ratio"
-]].rename(columns={
-    "year": "Year",
-    "total_deaths": "Total Deaths",
-    "mean_deaths": "Mean",
-    "median_deaths": "Median",
-    "min_deaths": "Min",
-    "max_deaths": "Max",
-    "male_ratio": "Male Ratio (%)",
-    "female_ratio": "Female Ratio (%)",
-    "other_ratio": "Other Ratio (%)"
-})
+    ax.set_title(str(year), fontsize=10)
+    ax.set_xlabel("Month")
+    ax.set_ylabel("Total Deaths")
+    ax.tick_params(axis='x', rotation=45)
 
-# Display styled table with caption
-summary_table.style.set_caption("Annual Summary of Deaths in Toronto Shelters (2007–2023)") \
-    .set_table_styles([
-        {"selector": "caption", "props": [("caption-side", "top"), ("font-size", "110%"), ("font-weight", "bold")]},
-        {"selector": "th", "props": [("text-align", "center")]},
-        {"selector": "td", "props": [("text-align", "center")]}
-    ]) \
-    .set_properties(**{"font-family": "Cambria"}) \
-    .format(na_rep="NA")
+# Hide any unused subplots
+for j in range(i + 1, len(axes)):
+    fig.delaxes(axes[j])
+
+# Add a main title
+fig.suptitle("Deaths in Toronto Shelters Over Time (2007.01–2024.08)",
+             fontsize=16, fontweight='bold', y=1.02)
+
+plt.tight_layout()
+plt.show()
+
 ```
